Article Title

### 2013-10-01 sections/Acquisition.tex
updated the Theory section.
redistributed text between files, cleaned up text files.

\subsection{Data acquisition}
Describe the two-photon experiment, refer to Fig.\;1

>>>>

\subsection*{Data acquisition}
We recorded the calcium activity  of dense populations of neurons in the supragranular layers in primary visual cortex of anesthetized mice using fast random-access 3D scanning two-photon microscopy \cite{Stosiek:2003,Reddy:2005}.  Numerious repetitions of full-field drifting gratings (Fig. 1A and 1B) were presented to the eye contralateral to the imaged site. This technique allowed recording from a large number (150--350) of cells in a small volume of cortical tissue ($200\times200\times100$ $\mu$m$^3$) in layers 2/3 and 4. Somatic calcium signals were deconvolved using  sparse nonnegative deconvolution \cite{Vogelstein:2010} (Fig.\;1C and 1D).  The average stimulus response was subtracted to remove stimulus covariance and the sample noise covariance matrix was computed (Fig.\;1E).

\input{sections/Figure1.tex}

### 2013-10-01 sections/Estimators.tex
redistributed text between files, cleaned up text files.

The covariance matrix estimate is then 
\begin{equation}
-\hat\Sigma_\rho^{-1} =\arg\min\limits_S \loss{-S^{-1},\hat\Sigma_0}  \quad \mbox{such that} \quad  \|S_\rho\|_0 = \rho
\end{equation}
where $S_\rho$ is a sparse matrix with $\rho$ non-zero coefficients. $\|S\|_0$ denotes the number of non-zero elements of matrix $S$.
Solving this approximation in this form is computationally challenging. The problem can be made computationally tractable by replacing the $L_0$ norm with the $L_1$ norm, which converts the problem into one of convex optimization \cite{Donoho:2000}.  The resulting algorithm, known as \emph{graphical lasso} or \emph{GLASSO}  \cite{Meinshausen:2006,Banerjee:2008,Friedman:2008}, produces the following estimator  
\begin{equation}
-\hat\Sigma_\lambda^{-1} = \arg\min\limits_S \loss{-S^{-1},\hat\Sigma_0} + \lambda \|S\|_1
\end{equation}
where $\lambda \|S\|_1$ is the $L_1$ norm of the matrix. \TODO{clarify}.

>>>>

The covariance matrix estimate is then 
\begin{equation}
-\hat\Sigma_\rho^{-1} =\arg\min\limits_S \loss{-S^{-1},\hat\Sigma_0}  \quad \mbox{such that} \quad  \|S_\rho\|_0 = \rho
\end{equation}
where $S_\rho$ is a sparse matrix with $\rho$ non-zero coefficients. $\|S\|_0$ denotes the number of non-zero elements of matrix $S$.
Solving this approximation in this form is computationally challenging. The problem can be made computationally tractable by replacing the $L_0$ norm with the $L_1$ norm, which converts the problem into one of convex optimization \cite{Donoho:2000}.  The resulting algorithm, known as \emph{graphical lasso} or \emph{GLASSO}  \cite{Meinshausen:2006,Yuan:2007,Banerjee:2008,Friedman:2008}, produces the following estimator  
\begin{equation}
-\hat\Sigma_\lambda^{-1} = \arg\min\limits_S \loss{-S^{-1},\hat\Sigma_0} + \lambda \|S\|_1
\end{equation}
where $\lambda \|S\|_1$ is the $L_1$ norm of the matrix. \TODO{clarify}.

### 2013-10-01 sections/Theory.tex
redistributed text between files, cleaned up text files.

\subsection*{Basic concepts}
Let $x(t) \in \mathcal X, t=1\ldots,n$ denote a sample of consecutive observations of population activity of $p$ neurons in time bins $t$.  
The activity of each neuron is denoted by the real-valued firing rate, thus  $\mathcal X = \mathbb R^{p\times 1}$.  
For example, in our application, the firing rate is estimated from somatic calcium fluorescence signals with the mean stimulus response subtracted, making it possible for elements of $x(t)$ to take on negative and non-integer values.

We do not assume that the observations $x(t)$ are independently distributed. Rather, the data generating process is assumed to be \emph{ergodic}, \emph{i.e.}\;described by a \emph{true distribution} $F: \mathcal X \mapsto  \mathbb [0, 1]$ (cumulative) over long periods of time such that the \emph{empirical distribution} $\hat F_n$ from a given sample will convege to $F$ with increasing sample size $n$.

Formally, the emprical distribution is defined as \TODO{this may be unecessary, but I leave it for now.}
\begin{equation}
\hat F_n(x) = \frac 1 n \sum\limits_{t=1}^n \mathbf{1}(x \ge x(t))
\end{equation}
where $\mathbf 1(x \ge x(t))$ is the indicator function which equals 1 when all elements of $x$ are greater than the corresponding elements of $x(t)$ and 0 otherwise. \TODO{KJ: Are you using cumulative distributions to avoid binning? DY: Yes. Many theorems such as Glivenko-Catelli Theorem and Skorohod's Representation are proven using  cumulative representation. The empirical distribution is nearly always represented in its cumulative form. Expressing convergence is requires binning or using cumulative; the latter is simpler.}

For an ergodic process, $\max\limits_{x\in\mathcal X} \left|\hat F_n(x) - F(x)\right| \to 0$ as $n$ increases.  \TODO{If you decide to leave this in, provide a reference to a standard book in the field.}

The true covariance matrix $\Sigma \in \Theta$ is defined as a function of $F$:
\begin{equation}
\Sigma = \int\limits_{z\in\mathcal X} (z - \mu)(z - \mu)^\T \dif F(z)
\quad\mbox{where}\;
\mu = \int\limits_{z\in\mathcal X} z \dif F(z)
\end{equation}
The domain $\Theta$ is the set of all positive-definite $p\times p$ matrices, which is a cone in $\mathbb R^{p\times(p+1)/2}$.

But the usual estimator of the true covariance matrix $\Sigma$ is  the sample covariance matrix:
\begin{equation}
\hat\Sigma_0 =  \frac 1 {n-c} \sum\limits_{t=1}^n (x(t) -\hat\mu) (x(t) - \hat\mu)^\T
\quad\mbox{where}\;
\hat\mu = \frac 1 n \sum\limits_{t=1}^n x(t)
\end{equation}
where $c$ is the sum of temporal correlations \TODO{there must be a better description of this} \TODO{Which?  I think the equation is OK.  Do you mean $c$?  This is standard, and again, I suggest just providing a reference.}. For independent observations, $c=1$;  $c>1$ when nearby samples are correlated; the value of $c$ can be estimated from the data.
\TODO{Under the Gaussian loss, $\mathcal L_g$ (\ref{eq:GaussLoss}) the scaling is irrelevant, so we will not spend any time talking about the scaling.  But it would be important if we used $\mathcal L_e$ (\ref{eq:MSE}).}

$\hat\Sigma_0$ is designed to be unbiased such that $\mathbb E\left[\hat\Sigma_0\right]=\Sigma$.
Here and througout, $\mathbb E[\cdot]$ denotes the expected value under the  unknown true distribution $F$. All variables with a hat (\emph{e.g.}\;$\hat \Sigma_0,\hat \mu$) are functions of the empirical distribution $\hat F_n$, which itself is a random variable.

%Nonlinear functions of an unbiased estimate can be biased: $\mathbb E\left[\hat\Sigma_0\right]=\Sigma \centernot\implies  E\left[\varphi(\hat\Sigma_0)\right]=\varphi(\Sigma)$.  For example, if $u_{\max}(S)$ is the largest eigevalue of square matrix $S$, then $\mathbb E\left[u_{\max}(\hat\Sigma_0)\right] > u_{\max}(\Sigma)$. Some covariance matrix estimators are designed to correct for the eigenspectrum bias at the cost  of adding bias to covariance coefficients \cite{Ledoit:2004}.

\subsection*{Loss and risk}
The optimization of a covariance matrix estimate $\hat\Sigma$ is performed with respect to a \emph{loss function} $\mathcal L(\hat\Sigma,\Sigma)$, which expresses the discrepancy between $\hat\Sigma$ and $\Sigma$ and attains its minimum when $\hat\Sigma=\Sigma$.  
Then \emph{excess loss}  
\begin{equation}
\ell(\hat\Sigma,\Sigma) = \mathcal L(\hat\Sigma,\Sigma)-\mathcal L(\Sigma,\Sigma)
\end{equation}
assumes zero at its minimum.

A particularly useful loss function is the mean squared error (MSE), which is proportional to the square of the Frobenius  norm $\|\cdot\|_F$ of the difference between the two matrices: 
\begin{equation}\label{eq:MSE}
\mathcal L_e(\hat\Sigma,\Sigma) =\frac 1 p \|\hat\Sigma-\Sigma\|_F^2 = \frac 1 p \Tr\left((\hat \Sigma-\Sigma)(\hat\Sigma-\Sigma)^\T\right)
\end{equation}
Since $\mathcal L_e(\Sigma,\Sigma) \equiv 0$, the MSE is its own excess loss: $\ell_e(\hat\Sigma,\Sigma) \equiv \mathcal L_e(\hat\Sigma,\Sigma)$.

The Gaussian loss function $\mathcal L_g$  arises from the theory of multivariate normal distributions. When observations are identically and independently distributed according to a multivariate normal distribution with zero means, the log likelihood of the covariance matrix $\Sigma$ with $\hat\Sigma = \frac 1 n \sum\limits_{t=1}^n x(t) x(t)^\T$ is  
\begin{equation}
L\left(\Sigma \mid \hat\Sigma\right) = -\frac n 2 \ln(2\pi) - \frac n 2 \ln \det \Sigma - \frac n 2 \Tr(\Sigma^{-1} \hat \Sigma)
\end{equation}
Then $\mathcal L_g$ is constructed by rescaling $L\left(\Sigma \mid \hat\Sigma\right)$ and dropping the constant term:
\begin{equation}\label{eq:GaussLoss}
\mathcal L_g(\hat\Sigma,\Sigma) 
=  -\frac 2 {pn} L\left(\hat\Sigma \mid \Sigma \right) - \frac 1 p \ln(2\pi) 
\equiv  \frac 1 p\left(\ln \det \hat \Sigma + \Tr(\hat \Sigma^{-1}\Sigma) \right) 
\end{equation}
\TODO{Note that the switch from $L\left(\Sigma \mid \hat\Sigma \right)$ to $L\left(\hat\Sigma \mid \Sigma \right)$ is on purpose, which will become clear in the next few steps.}
The corresponding excess loss 
\begin{equation}
\ell_g(\hat\Sigma,\Sigma) = \mathcal L_g(\hat\Sigma,\Sigma) - \mathcal L_g(\Sigma,\Sigma)  
= \frac 1 p \left(-\ln \det (\hat \Sigma^{-1} \Sigma) + \Tr(\hat \Sigma^{-1}\Sigma)\right) - 1
\end{equation}
is known as \emph{entropy loss} \cite{James:1961}. \TODO{The order of the operands is reversed. Check that it's okay.}

Despite the fact that entropy loss is derived from normal theory, the choice of a loss function is not equivalent to assuming a specific form of $F$. The loss function measures the discrepancy between distribution parameters rather than the distance between the distributions themselves. \TODO{explain Bregman divergence?}

The expected value of excess loss is the \emph{estimator risk}:
\begin{equation}\label{eq:risk}
r = \mathbb E\left[\ell(\hat\Sigma,\Sigma)\right]
\end{equation}
The estimator risk is the primary quality criterion for covariance estimation. Estimator $\hat\Sigma_a$ is considered more \emph{efficient} than estimator $\hat\Sigma_b$ if $\hat\Sigma_a$ has lower risk for the given sample than $\hat\Sigma_b$.   \TODO{explain that the risk depends not only on $\hat\Sigma$ but also on the distribution of $\Sigma$ in the specific domain.}

\subsection*{Cross validation}
In practice, the true covariance matrix $\Sigma$ is not accessible and the estimator risk $r$ must be estimated from data. This requires a separate \emph{validation} empirical distribution $\hat F_m^\prime$ of $m$ observations sampled from $F$ independently of the \emph{training} distribution $\hat F_n$. 
Then let  $\hat \Sigma_0^\prime$ be the sample covariance obtained from $\hat F_m^\prime$.

Then the \emph{empirical loss} of $\hat\Sigma$ is $\mathcal L(\hat\Sigma,\hat\Sigma_0)$ and its expectation is the \emph{empirical risk}  
\begin{equation}
\hat r = \mathbb E\left[\mathcal L(\hat\Sigma,\hat\Sigma_0^\prime)\right]
\end{equation}

 The two loss functions $\mathcal L_e(\hat\Sigma,\Sigma)$ and $\mathcal L_g(\hat\Sigma,\Sigma)$ are particularly suitable for risk estimation thanks to their linearity with respect to $\Sigma$ in the sense that 
\begin{equation}
\mathcal L\left(\hat\Sigma,\alpha S_1 + (1-\alpha)S_2\right) 
\equiv 
\alpha\mathcal L(\hat \Sigma,S_1) + (1-\alpha)\mathcal L(\hat \Sigma,S_2)
\end{equation}
which allows bringing the expectation inside the empirical loss function:
\begin{equation}
\hat r = 
\mathbb E\left[ \mathcal L(\hat\Sigma, \hat\Sigma_0^\prime) \right] 
=
\mathbb E\left[ \mathcal L\left(\hat\Sigma, \mathbb E\left[\hat\Sigma_0^\prime\right]\right) \right] 
=
\mathbb E\left[ \mathcal L(\hat\Sigma, \Sigma) \right] 
= 
r + \mathcal L(\Sigma,\Sigma)
\end{equation}
This means, that the empirical loss $\mathcal L(\hat\Sigma,\hat\Sigma_0^\prime)$ is an unbiased estimate of the estimator risk $r$ (up to a constant offset): minimization of the empirical loss implies minimization of the true estimator risk.

Recording a separate independent validation sample is usually not sensible.  Instead, the recorded data are split into a training sample and validation sample. In $K$-fold \emph{cross-validation} the data are split into $K$ roughly equal-sized subsets. Each subset then successively serve as the testing sample while the remainder of the data serves as the training sample.  This procedure produces $K$ estimates of the  risk, which are then averaged together to produce a better averaged estimate.

\subsection*{Bias/variance decomposition}
The estimator risk (\ref{eq:risk}) can be decomposed into \emph{approximation error} or \emph{bias}   
\begin{equation}
b^2 = \ell \left( \mathbb E[\hat \Sigma],\Sigma\right)
\end{equation}
and \emph{estimation error} or \emph{variance}
\begin{equation}
\varepsilon^2 = \mathbb E \left[ \ell\left(\hat \Sigma, 
\mathbb E[\hat \Sigma]\right) \right]
\end{equation}

Under the MSE loss function $\mathcal L_e$ (\ref{eq:MSE}), the decomposition is the simple sum $r = b^2 + \varepsilon^2$. 
Under other loss functions, $r$ is an increasing function of both $b^2$ and $\varepsilon^2$ although not generally a simple sum.  
\TODO{As far as I know, the terms `bias' and `variance' are only applicable under a quadratic loss function  like $\mathcal L_e$.  Under other loss functions, we may need to use approximation/estimation error.}

The sample covariance matrix $\hat\Sigma_0$ has zero bias but high variance. Other estimators may be biased but have lower estimation error and potentially lower risk.

>>>>

\subsection*{The sample covariance matrix}
We aim to estimate the true covariance matrix $\Sigma = \E{(x-\mu)(x-\mu)^\T}$ of the instantaneous activity vector $x$ of a population of $p$ neurons. Here $\E{\cdot}$ denotes expectation \TODO{for the true data generating process} and $x$ is the $p\times 1$ vector of real-valued instantaneous firing rates discretized into bins of duration $\Delta t$ and $\mu = \E{x}$.

For more rigorous notation, definitions, and derivations, see Appendix.

The usual estimator of $\Sigma$ is the sample covariance matrix
\begin{equation}
\hat \Sigma_0 = \frac 1 \nu \sum\limits_{t=1}^n (x(t)-\mu)(x(t)-\mu)^\T 
\end{equation}
where $x(t),\;t=1,\ldots,n$ are sequential observations of population activity inferred from calcium signals; $\nu$ is the number of degrees of freedom. For independent observations $\nu=n-1$ because estimation of the mean $\mu$ accounts for one degree of freedom. When observations are correlated, as is the case with calcium signals, $\nu < n-1$ and may be estimated from the signal.

The sample covariance matrix is constructed to be unbiased, such that $\E{\hat\Sigma_0} = \Sigma$.

\subsection*{Evaluation of covariance matrix estimates}
The quality of a covariance matrix estimate $\hat\Sigma$ is measured by a real-valued \emph{loss function} $\loss{\hat\Sigma,\Sigma}$.  The loss function quantifies the deviation of $\hat\Sigma$ from $\Sigma$ and attains its minimum  when $\hat\Sigma = \Sigma$.

For the purposes of this study, we adopted the \emph{negative normal log-likelihood loss} function:
\begin{equation}\label{eq:loss}
\loss{\hat\Sigma,\Sigma} = \frac 1 p\left[\ln \det \hat \Sigma + \Tr(\hat \Sigma^{-1}\Sigma)\right]
\end{equation}
This choice is motivated by mathematical convenience. Other popular choices for the loss function are the Frobenius norm of the difference $\hat\Sigma-\Sigma$, Stein's entropy loss, and quadratic loss \cite{James:1961,Ledoit:2004,Schafer:2005,Fan:2008}.  We assert that the main conclusions of our study are unlikely to change drastically other well behaved loss functions.

The aim of our project is produce covariance matrix estimates that minimize the expected loss.  The expected loss of an estimator is known as its \emph{risk}: 
\begin{equation}
r = \E{\loss{\hat\Sigma, \Sigma}}
\end{equation}

In practice, the true value $\Sigma$ is not accessible and estimators' risks must be estimated from the data.  This may be accomplished through validation: 
Let $\hat\Sigma_0^\prime$ denote a sample covariance matrix measured from an independent sample that was not used included in the computation of $\hat\Sigma$. Then \emph{empirical loss} is 
\begin{equation}
\hat \ell = \loss{\hat\Sigma,\hat\Sigma_0^\prime}
\end{equation}
 and its expected value or \emph{empirical risk} is
 \begin{equation}\label{eq:empiricalRisk}
 \hat r = \E{\hat\ell} = \mathbb E_{\hat\Sigma} \left[ \mathbb E_{\hat\Sigma_0^\prime} \left[ {\loss{\hat\Sigma,\hat\Sigma_0^\prime}} \right] \right]
 \end{equation}
 Because the chosen loss function is linear in its second argument in the sense that
 \begin{equation}
 \loss{\hat\Sigma,X_1} + \loss{\hat\Sigma,X_2} \equiv \loss{\hat\Sigma,X_1+X_2}
 \end{equation}
 the expection on the second argument in Eq.~\ref{eq:empiricalRisk} may be taken inside the loss function:
 \begin{equation}
 \hat r = \E{\loss{\hat\Sigma,\E{\hat\Sigma_0^\prime}}}  = \E{\loss{\hat\Sigma,\Sigma}} = r
 \end{equation}
 Thus the empirical loss $\loss{\hat \Sigma,\hat \Sigma_0^\prime}$ serves as an unbiased estimate of risk $r$.

 Because the loss function is equivalent to the negative normal log likelihood, the above derivation led us to the familiar criterion that the optimal covariance matrix estimator is one that consistently maximizes the normal log likelihood of the validation dataset.

 \subsection*{Regularization}
 Under many loss functions\footnote{
 The strict equality in Eq.~\ref{eq:bias-variance} does not hold under the loss function in Eq.~\ref{eq:loss}. However, the equality does hold for its close cousin, Stein's \emph{entropy loss},  which only differs by the order of its arguments and a constant offset: $\mathcal L_s(\hat\Sigma,\Sigma) \equiv \loss{\Sigma,\hat\Sigma} - \loss{\Sigma,\Sigma}$. This defficiency presents no difficulty because we minimize the risk directly, without assessing the two error components individually. The bias-variance decomposition is presented here to motivate the use of regularization.}, 
 the estimator's risk can be decomposed as the sum
 \begin{equation}\label{eq:bias-variance}
 r = b + \varepsilon
 \end{equation}
 of \emph{approximation error} (``bias'' or systematic error)
 \begin{equation}
 b = \loss{\bar\Sigma,\Sigma}
 \end{equation}
 and \emph{estimation error} (``variance'') 
 \begin{equation}
 \varepsilon = \E{\loss{\hat\Sigma, \bar\Sigma}}
 \end{equation}
 where $\bar\Sigma = \E{\hat\Sigma}$ is the expected value of the estimate.

 The unbiased estimator $\hat\Sigma_0$ makes $\bar\Sigma=\Sigma$ and thereby minimizes the approximation error, but may be excessively susceptible to sample noise and result in high estimation error.

 The estimator risk can be reduced by \emph{regularization}. Regularization is the deliberate biasing (\emph{``shrinkage''}) of the estimate toward a low-dimensional, less variable \emph{target estimate} \cite{Bickel:2006,Ledoit:2004}. A regularized estimator solves the bias-variance tradeoff to produce a biased but less variable estimates aiming to minimize the estimator's risk.  Various regularization schemes focus on the dimensionality reduction part \TODO{rephrase} by selecting the optimal target estimate from a family of estimates with reduced dimensionality \cite{findit}.  Other estimators only shrink the sample covariance matrix toward a single target estimator \cite{Schafer:2005}. Yet other regularizers effectively combine shrinkage and dimensionality reduction \cite{findit}.
